Appendix A - Alternative DSDM formulations

### 2019-10-30 manuscript/appendix_a.Rmd
add notation and description for IPMs

@@

Each element of the stage-dispersal matrices $\mathbf{D}_j$ give the probability $D_{j,i,i'}$ that an individual of  stage $j$ in site $i$ will move to site $i'$ in this time-step. $D_{j,i,i'}$ is computed using a dispersal kernel (an exponential kernel in this example) with stage-specific dispersal ability parameter $\tau_j$, standardised to sum to 1 over the study area to ensure $D_{j,i,i'}$ is a valid probability.

@@

In practice, the matrix $\mathbf{A}^*$ can be very sparse: the proportion of non-zero elements is $(M + N - 1) / (MN)$; as N increases, the proportion of non-zero approaches $1/M$. This sparsity can be increased further by setting an upper dispersal limit, beyond which $d_{j,i,i'}$ is set to zero, or by limiting the number of stages that can disperse. It is therefore computationally advantageous to compute the dominant eigenvector using sparse matrix methods, e.g. by iteration using sparse matrix multiplication, as in the biotic interaction model.

@@

\begin{align}
\Lambda_i &= \alpha \mathbf{v}_i / \textstyle\sum_{i=1}^{N}{\mathbf{v}_{i}} \\ \nonumber
\mathbf{v}_i &= \textstyle\sum_{j=1}^{M}{\mathbf{v}^*_{j+(i-1)}} \\ \nonumber
\mathbf{v}^* &= eigvec(\mathbf{A}^*) \\ \nonumber
\mathbf{A}^* &= \mathbf{P} \mathbb{A} \mathbf{P}^T \mathbb{D} \\ \nonumber
\mathbb{A} = \begin{bmatrix} \mathbf{A}_1 & & \dots & & 0 \\
& \ddots & & & \\
\vdots & & \mathbf{A}_i & & \vdots \\
& & & \ddots & \\
0 & & \dots & & \mathbf{A}_N \end{bmatrix}
&\,\, \mathbb{D} = \begin{bmatrix} \mathbf{D}_1 & & \dots & & 0 \\
& \ddots & & & \\
\vdots & & \mathbf{D}_j & & \vdots \\
& & & \ddots & \\
0 & & \dots & & \mathbf{D}_M \end{bmatrix} \\ \nonumber
\mathbf{D}_{j,i,i'} &= d_{j,i,i'} / \textstyle\sum_{i=1}^Nd_{j,i,i'} \\ \nonumber
d_{j,i,i'} &=  e^{-\tau_j D_{i,i'}}
\end{align}
where $D$ is an $N$x$N$  dense, symmetric matrix of Euclidean distances between all sites.

@@

\begin{align}
\mathbf{n}_{j,t+1} &= \mathbf{D}_j \mathbf{n}^*_{j,t+1} \\ \nonumber
\mathbf{n}^*_{i,t+1} &= \mathbf{A}_{i,t} \mathbf{n}_{i,t} \\ \nonumber
\mathbf{D}_{j,i,i'} &= d_{j,i,i'} / \textstyle\sum_{i=1}^Nd_{j,i,i'} \\ \nonumber
d_{j,i,i'} &=  e^{-\tau_j D_{i,i'}}
\end{align}

@@

<to do, don't package up integration code into greta.dynamics yet, show it warts-and-all for now>

>>>>

\newcommand{\smallunderscore}{\textscale{.5}{\textunderscore}}

@@

Each element of the stage-dispersal matrices $\mathbf{D}_j$ give the probability $D_{j,i',i}$ that an individual of  stage $j$ in site $i$ will move to site $i'$ in this time-step. $D_{j,i',i}$ is computed using a dispersal kernel (an exponential kernel in this example) with stage-specific dispersal ability parameter $\tau_j$, standardised to sum to 1 over the study area to ensure $D_{j,i',i}$ is a valid probability.

@@

In practice, the matrix $\mathbf{A}^*$ can be very sparse: the proportion of non-zero elements is $(M + N - 1) / (MN)$; as N increases, the proportion of non-zero approaches $1/M$. This sparsity can be increased further by setting an upper dispersal limit, beyond which $d_{j,i',i}$ is set to zero, or by limiting the number of stages that can disperse. It is therefore computationally advantageous to compute the dominant eigenvector using sparse matrix methods, e.g. by iteration using sparse matrix multiplication, as in the biotic interaction model.

@@

\begin{align}
\Lambda_i &= \alpha \mathbf{v}_i / \textstyle\sum_{i=1}^{N}{\mathbf{v}_{i}} \\ \nonumber
\mathbf{v}_i &= \textstyle\sum_{j=1}^{M}{\mathbf{v}^*_{j+(i-1)}} \\ \nonumber
\mathbf{v}^* &= eigvec(\mathbf{A}^*) \\ \nonumber
\mathbf{A}^* &= \mathbf{P} \mathbb{A} \mathbf{P}^T \mathbb{D} \\ \nonumber
\mathbb{A} = \begin{bmatrix} \mathbf{A}_1 & & \dots & & 0 \\
& \ddots & & & \\
\vdots & & \mathbf{A}_i & & \vdots \\
& & & \ddots & \\
0 & & \dots & & \mathbf{A}_N \end{bmatrix}
&\,\, \mathbb{D} = \begin{bmatrix} \mathbf{D}_1 & & \dots & & 0 \\
& \ddots & & & \\
\vdots & & \mathbf{D}_j & & \vdots \\
& & & \ddots & \\
0 & & \dots & & \mathbf{D}_M \end{bmatrix} \\ \nonumber
\mathbf{D}_{j,i',i} &= d_{j,i',i} / \textstyle\sum_{i'=1}^Nd_{j,i',i} \\ \nonumber
d_{j,i',i} &=  e^{-\tau_j D_{i',i}}
\end{align}
where $D$ is an $N$x$N$  dense, symmetric matrix of Euclidean distances between all sites.

@@

\begin{align}
\mathbf{n}_{j,t+1} &= \mathbf{D}_j \mathbf{n}^*_{j,t+1} \\ \nonumber
\mathbf{n}^*_{i,t+1} &= \mathbf{A}_{i,t} \mathbf{n}_{i,t} \\ \nonumber
\mathbf{D}_{j,i',i} &= d_{j,i',i} / \textstyle\sum_{i=1}^Nd_{j,i',i} \\ \nonumber
d_{j,i',i} &=  e^{-\tau_j D_{i',i}}
\end{align}

@@

For species with complex demography and populations that are better represented as size-structureed rather than stage-structured DSDMs can be formed by replacing the MPM with an integral projection model (IPM). As an example, this section presents the equivalent code and notation for the complex, environmentally-explicit spatial IPM for *Protea repens* presented in @merowUsingIntegralProjection2014 and studied as a DSDM in the manuscript. This model is particularly complex - having 8 vital rate submodels, and requiring interpolation of survival rates betwen seedlings and adults.

Whereas matrix population models consider populations with a discrete population distribution over different stages in the model, integral projection models consider populations via a continuous population distribution over individual sizes. The dynamics of a population at site $i$ are therefore modelled as $\mathbf{n}_{i,t+1}(z') = \int \mathbf{K}_{i}(z'|z)\mathbf{n}_{i,t}(z) \, dz$, where $\mathbf{n}_{i,t}(z)$ is an abundance *function* (at site $i$ and time $t$) of sizes $z$, giving the number of individuals of with size around $z$, and $\mathbf{K}_{i}(z'|z)$ is a continuous transition *kernel* at site $i$ mapping these abundance functions between time points. Whereas a matrix population model is described by a matrix-vector multiplication (which sums products between elements of the matrix and vector), the analagous operaiton in this continuous representation is an integration over a convolution. In practice though, this integration and convolution is commonly approximated by discretizing and normalising the kernel $\mathbf{K}_{i}(z'|z)$ to yield a matrix $\mathbf{A}_i$ with a sufficiently large number of size bins $j$ which can be analysed in the same way as a matrix population model. We can therefore compute such a matrix and include it in the DSDM framework.

Whereas MPM-based DSDMs have a single environmental regression submodel for each vital rate, in the IPM these are replaced with submodels that combine the effects of both the environment and individual size. These functions must therefore be manipulated to create arrays (over sites and size bins) that are then combined and normalised to yield an approximation to the IPM. The discretized notation and code to implement these models is therefore somewhat more involved than for MPMs.

\begin{align}
\mathbf{A}_i &= \mathbf{F}_i + \mathbf{P}_i \\ \nonumber
\mathbf{F}_{i,j',j} &= p_{\text{flower}_{j}} f_{\text{seedheads}_i} f_{\text{seeds}} p_{\text{germ}} f^*_{\text{offsize}_{i,j'}} \\ \nonumber
\mathbf{P}_{i,j',j} &= p_{\text{surv}_{i,j}} f^*_{\text{grow}_{i,j',j}} \\ \nonumber
f^*_{\text{offsize}_{i,j'}} &= f_{\text{offsize}_{i,j'}} / \textstyle\sum_{j'=1}^{M}{f_{\text{offsize}_{i,j'}}} \\ \nonumber
f^*_{\text{nextsize}_{i,j',j}} &= f_{\text{nextsize}_{i,j',j}} / \textstyle\sum_{j'=1}^{M}{f_{\text{nextsize}_{i,j',j}}} \\ \nonumber
logit(p_{\text{flower}}) &= \alpha_{\text{flower}} + z \gamma_{\text{flower}} \\ \nonumber
log(f_{\text{seedheads}}) &= \mathbf{X}_{\text{seedheads}} \beta_{\text{seedheads}} \\ \nonumber
p_{\text{surv}_{i,j}} &= w_j p_{\text{surv-adult}_{i,j}} + (1 - w_j) p_{\text{surv-seedling}_i}  \\ \nonumber
logit(p_{\text{surv-adult}_{i,j}}) &= \mathbf{X}_{\text{surv}_i} \beta_{\text{surv-adult}} + z_j \gamma_{\text{surv-adult}} \\ \nonumber
logit(p_{\text{surv-seedling}_i}) &= \mathbf{X}_{\text{surv}_i} \beta_{\text{surv-seedling}} + z_j \gamma_{\text{surv-seedling}} \\ \nonumber
f_{\text{offsize}_{i,j'}} &= \pi (z_{j'} | \mu_{\text{offsize}_{i,j'}}, \sigma_{\text{offsize}}^2) \\ \nonumber
\mu_{\text{offsize}_{i,j'}} &= \mathbf{X}_{\text{offsize}_i} \beta_{\text{offsize}} \\ \nonumber
f_{\text{nextsize}_{i,j',j}} &= \pi (z_{j'} | z_{j} + \mu_{\text{grow}_{i,j}}, \sigma_{\text{grow}}^2) \\ \nonumber
\mu_{\text{grow}_{i,j}} &= \mathbf{X}_{\text{grow}_i} \beta_{\text{grow}} + z_{j} \gamma_{\text{grow}} \\ \nonumber
\end{align}

where $z$ is a vector of sizes of length $M$; $\mathbf{X}_{\text{seedheads}}$, $\mathbf{X}_{\text{surv}}$, $\mathbf{X}_{\text{offsize}}$, $\mathbf{X}_{\text{grow}}$, $\beta_{\text{seedheads}}$, $\beta_{\text{surv-adult}}$, $\beta_{\text{surv-seedling}}$, $\beta_{\text{offsize}}$, and $\beta_{\text{grow}}$ are design matrices of environmental covariates (and intercept columns) and vectors of regression coefficients for the environmental components of log-seedhead production, logit- adult and seedling survival probabilities, mean offspring size and mean growth rates; $\alpha_{\text{flower}}$, $\gamma_{\text{flower}}$, $\gamma_{\text{surv-adult}}$, $\gamma_{\text{surv-seedling}}$, and $\gamma_{\text{grow}}$ are scalar intercepts and regression coefficients for the effect of size on the logit-probabilities of flowering, adult survival, seedling survival, and on the growth rate; $f_{\text{seeds}}$ and $p_{\text{germ}}$ are scalar parameters for the average number of seeds per seedhead and the probability of germination; $w$ is a vector of weights of length $M$ linearly interpolating size-specific survival between adult and seedling survival probabilities; $\sigma_{\text{offsize}}$ and $\sigma_{\text{grow}}$ are positive scalar parameters giving the standard deviations of offspring sizes and growth rates; and $\pi (x | \mu, \sigma ^ 2)$ denotes the density of univariate normal distribution with parameters $\mu$ and $\sigma^2$ evaluated at $x$.

The below code requires some custom functions and a custom greta operation to perform the integration steps and build the kernels efficiently across 3D arrays representing kernels across multiple sites, and to carry out the interpolation of survival rates between adults and seedlings that is implemented in @merowUsingIntegralProjection2014.

- `matrix_dim`: a positive integer giving $M$ - the dimension of the matrix used to integrate the IPM kernel
- `X_offsize`: a $N$x$K_{\text{offsize}}$ design matrix for the offspring size environmental sub-model
- `X_surv`: a $N$x$K_{\text{surv}}$ design matrix for the survival environmental sub-models
- `X_grow`: a $N$x$K_{\text{grow}}$ design matrix for the growth environmental sub-model
- `X_seedheads`: a $N$x$K_{\text{seedhead}}$ design matrix for the seedhead environmental sub-model
- `beta_offsize`: a length $K_{\text{offsize}}$ vector of environmental regression coefficients for offspring size
- `beta_surv_seedling`: a length $K_{\text{surv}}$ vector of environmental regression coefficients for logit-seedling-survival
- `beta_surv_adult`: a length $K_{\text{surv}}$ vector of environmental regression coefficients for logit-adult-survival
- `beta_grow`: a length $K_{\text{grow}}$ vector of environmental regression coefficients for growth
- `beta_seedheads`: a length $K_{\text{seedhead}}$ vector of environmental regression coefficients for log-seadhead production
- `gamma_surv_seedling`: a scalar coefficient for the effect of size on logit-seedling-survival
- `gamma_surv_adult`: a scalar coefficient for the effect of size on logit-adult-survival
- `gamma_grow`: a scalar coefficient for the effect of size on growth
- `gamma_flower`: a scalar coefficient for the effect of size on logit-flowering-probability
- `germination`: a scalar parameter in $(0,1)$ for germination-probability
- `seeds`: a positive scalar parameter for seed-production
- `alpha_flower`: a scalar intercept parameter for on logit-flowering-probability
- `offsize_sd`: a positive scalar parameter for the standard deviation of the offspring size distribution
- `grow_sd`: a positive scalar parameter for the standard deviation of the growth distribution

```{r ipms}
# normalise a matrix to sum to one rowwise *in exponential space*. Ie. return y
# where: exp(y[i, ]) = exp(x[i, ]) / sum(exp(x[i, ]))
log_normalise <- function (x) {
  op <- greta::.internals$nodes$constructors$op
  op("log_normalise", x, tf_operation = "tf_log_normalise")
}

# tensorflow code for log normalisation
tf_log_normalise <- function (x) {
  sums <- tensorflow::tf$reduce_logsumexp(x, axis = 2L, keepdims = TRUE)
  x - sums
}

# multiply matrix B by each slice of the 3D array A in 3rd dimension
sweep_prod_3d <- function(A, B) {
  n <- dim(A)[1]
  m <- dim(A)[2]
  Al <- aperm(A, c(2, 3, 1))
  dim(Al) <- c(m, n * m)
  Bl <- t(B)
  dim(Bl) <- n * m
  C <- sweep(Al, 2, Bl, "*")
  dim(C) <- c(m, m, n)
  aperm(C, c(3, 1, 2))
}

# create 3D kernel array for Gaussian transitions (working with the log density
# for computational stability)
distribution_kernel <- function (mean_mat, sd, sizes, bin_width) {
  var <- sd ^ 2
  log_constant <- -0.5 * log(2 * pi) - log(sd)
  diff <- kronecker(sizes, mean_mat, FUN = "-")
  log_density <- log(bin_width) + log_constant - (diff ^ 2) / (2 * var)
  log_density <- log_normalise(log_density)
  dim(log_density) <- c(dim(mean_mat), ncol(mean_mat))
  log_density <- aperm(log_density, c(1, 3, 2))
  exp(log_density)
}

interpolate_survival <- function (adult_surv, seedling_surv, sizes) {
  range <- 0.728 - 0.182
  weights <- 0.728 / range  - (sizes / range)
  weights[sizes > 0.728] <- 0
  weights[sizes <= 0.182] <- 1
  weighted_seedling_surv <- sweep(seedling_surv, 2, weights, FUN = "*")
  weighted_adult_surv <- sweep(adult_surv, 2, 1 - weights, FUN = "*")
  weighted_seedling_surv + weighted_adult_surv
}

min_size <- 0
max_size <- 5
bounds <- min_size + (0:matrix_dim) * (max_size - min_size) / matrix_dim
sizes <- 0.5 * (bounds[-(matrix_dim + 1)] + bounds[-1])
bin_width <- sizes[2] - sizes[1]

# environmental components of sub models
offsize <- X_offsize %*% beta_offsize
grow <- X_grow %*% beta_grow
env_surv_seedling <- X_surv %*% beta_surv_seedling
env_surv_adult <- X_surv %*% beta_surv_adult
env_seedheads <- X_seedheads %*% beta_seedheads

# size components of submodels
flower <- ilogit(alpha_flower + sizes * gamma_flower)
size_surv_seedling <- sizes * gamma_surv_seedling
size_surv_adult <- sizes * gamma_surv_adult
size_seedheads <- sizes * gamma_seedheads

# vital rates at all sites
seedheads <- exp(kronecker(size_seedheads, env_seedheads))
surv_seedling <- ilogit(kronecker(size_surv_seedling, env_surv_seedling))
surv_adult <- ilogit(kronecker(size_surv_adult, env_surv_adult))

# kernelise growth * survival and recruitment * offspring size, and combine
survival <- interpolate_survival(surv_adult, surv_seedling, sizes)
next_size_density <- distribution_kernel(sizes + grow, grow_sd, sizes, bin_width)
P <- sweep_prod_3d(next_size_density, survival)
recruits <- flower * seedheads * seeds * germ
offsize_density <- distribution_kernel(offsize, offsize_sd, sizes, bin_width)
F <- sweep_prod_3d(offsize_density, recruits)
A <- P + F
```
